---
title: "The Scaling Behaviour of an MST-Based Stringy Scagnostic"
subtitle: "Understanding Sample-Size Dependence Through MST Geometry and Null Calibration"
author: "Tina Rashid Jafari"
date: 2026-10-01
bibliography: references.bib
keywords:
  - scagnostics
  - projection pursuit
  - minimum spanning tree
  - scaling behaviour
  - weighted diameter
  - null calibration
  - geometric statistics
  - high-dimensional visualization
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
  pdf:
    number-sections: true
    keep-tex: true
execute:
  echo: false
  warning: false
  message: false
---

```{r}
#| label: setup
library(cassowaryr)
library(igraph)
library(ggplot2)
library(patchwork)
library(knitr)
library(readr)
library(dplyr)
library(purrr)
library(tidyr)
library(scales)

col_weighted <- "#0072B2"
col_hop      <- "#D55E00"
col_mean     <- "#0072B2"
col_theory   <- "#CC79A7"
col_null     <- "#009E73"
col_grey     <- "grey72"

paper_theme <- theme_bw(base_size = 11) +
  theme(
    aspect.ratio = 1,
    panel.grid.minor = element_blank(),
    strip.background = element_rect(fill = "grey96", colour = "grey70"),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey30"),
    legend.position = "bottom"
  )
```


# Abstract

Scagnostics summarize geometric features of bivariate point clouds using graph-based quantities developed from Tukey's scatterplot-diagnostic ideas [@tukey1985graphics; @wilkinson2005graph; @wilkinson2008scagdist]. They are attractive candidates for projection pursuit because projection pursuit searches for informative low-dimensional views by optimizing an index over projections [@friedman1974projection; @huber1985projection; @friedman1987exploratory; @cook1995grandtour]. However, a projection-pursuit index should respond primarily to structure rather than to sample size alone; comparable scales and distributions were explicitly identified as desirable properties for scagnostic measures by @wilkinson2008scagdist. This paper investigates the finite-sample scaling behaviour of the MST-based `stringy05` scagnostic,

$$
S_n=\frac{D_n}{L_n},
$$

where $L_n$ is the total Euclidean length of the minimum spanning tree (MST) and $D_n$ is its weighted diameter. Although both quantities are computed from the same tree, they capture fundamentally different aspects of its geometry: $L_n$ accumulates length across all MST edges, whereas $D_n$ records only the accumulated Euclidean length of the longest path.

We use theoretical results for Euclidean MSTs together with simulation experiments to determine how these two components contribute to the sample-size dependence of Stringy. For bivariate uniform data, the denominator follows the theoretically established $\sqrt n$ scaling of total Euclidean MST length [@steele1988growth]. The numerator is more difficult: no directly applicable theoretical growth rate is available for the weighted Euclidean MST diameter. Empirically, it is not well described by either the longest-edge scaling from Euclidean MST theory [@penrose1997longest] or the $n^{1/3}$ graph-diameter benchmark obtained for a different random-MST model [@addarioberry2006diameter]. Among the candidate scaling functions examined, $\log n$ provides the strongest stabilization of the mean weighted diameter over the larger simulated sample sizes, although this is an empirical finite-sample result rather than an asymptotic claim.

Using these separate scalings yields

$$
\widetilde S_n
=
S_n\frac{\sqrt n}{\log n},
$$

which substantially reduces the trend in the centre of the distribution but introduces two important limitations: it does not fully stabilize the distribution and it destroys the natural $[0,1]$ bound of the original scagnostic. We also show that weighted and hop-based versions of Stringy are not interchangeable; they can select different MST paths because they measure accumulated geometric length and edge count, respectively.

These findings suggest that deterministic scaling is most useful for explaining *why* Stringy changes with sample size, rather than for redefining the index itself. We therefore motivate a complementary statistical solution: retain the original bounded Stringy measure and calibrate it relative to its sample-size-specific null distribution. In particular, we examine models for the upper null quantiles and outline a bounded null-calibrated transformation for projection pursuit. This separates geometric understanding from statistical interpretation and provides a principled route toward using Stringy as a projection-pursuit index.

# Introduction

Projection pursuit is built on a simple idea: if a high-dimensional dataset contains an informative low-dimensional view, an index should be able to recognize it. The method searches over projections and assigns high values to views displaying the type of structure of interest [@friedman1974projection; @huber1985projection; @friedman1987exploratory]. Guided and grand tours later connected this optimization perspective with dynamic visualization, allowing analysts to move through the space of projections while using an index to direct attention toward potentially informative views [@cook1995grandtour; @wickham2011tourr].


This seemingly simple task places strong demands on the index. A projection-pursuit index should respond to the structure present in a projection, but it should not change substantially because of features unrelated to that structure. Sample size is one such feature. If the same underlying distribution receives systematically different index values as the number of observations changes, then an optimizer may be responding partly to the finite-sample behaviour of the index rather than to the geometry of the data. For projection pursuit, this distinction is fundamental: an index is useful only if a high value can be interpreted as evidence of the target structure rather than as a consequence of how many points happen to be present.

Scagnostics are particularly appealing in this setting because they were designed to quantify visual characteristics of scatterplots. The idea has its roots in Tukey's broader program of exploratory data analysis and graphical methods [@tukey1975picturing; @tukey1977eda], and more specifically in Tukey and Tukey's notion of scatterplot diagnostics [@tukey1985graphics]. The modern graph-theoretic formulation represents a bivariate point cloud using structures such as the minimum spanning tree (MST), convex hull, and alpha hull, and then constructs numerical measures of properties such as stringy, sparsity, clumpy, convexity, and outlying [@wilkinson2005graph; @wilkinson2006highdim; @wilkinson2008scagdist].

Because scagnostics translate visual structure into numerical indices, they are natural candidates for high-dimensional exploration and projection pursuit. They have been used to guide searches through high-dimensional projections and evolving multivariate data [@wilkinson2006highdim; @anand2012random; @dang2013timeseer; @dang2014transforming]. Their statistical behaviour, however, matters just as much as their visual interpretation. Previous work has considered the distributions of scagnostic measures, the desirability of comparable scales, transformations intended to expose hidden structure, and sensitivity to preprocessing choices such as binning [@wilkinson2008scagdist; @dang2014transforming; @wang2020robustscag].


This paper focuses on one particular scagnostic, `stringy05`, whose definition is based entirely on the geometry of the MST. For a Euclidean MST $T_n$ constructed from $n$ bivariate observations, Stringy is defined as

$$
S_n=\frac{D_n}{L_n},
$$

where

$$
L_n=\sum_{e\in T_n}|e|
$$

is the total Euclidean length of the tree and

$$
D_n=\max_{u,v\in T_n} d_{T_n}(u,v)
$$


is the weighted graph diameter, namely the accumulated Euclidean length of the longest path through the MST.

The interpretation is intuitive. If the MST is almost one long unbranched path, most of the total tree length lies along its diameter, so $D_n$ is close to $L_n$ and $S_n$ approaches one. If the tree branches extensively, no single path contains most of the total length and Stringy becomes smaller [@wilkinson2005graph; @wilkinson2008scagdist].

The difficulty is hidden inside this apparently simple ratio. The numerator and denominator are not the same kind of random geometric object. The denominator is additive: it sums the lengths of all $n-1$ MST edges. The numerator is extremal: it first identifies the longest weighted path in the tree and then sums only the edge lengths along that path. There is therefore no reason to expect the two quantities to scale identically with sample size.

This distinction becomes practically important because the null behaviour of Stringy changes with $n$. In simulations of unstructured data, the upper tail of the Stringy distribution decreases as sample size increases. Consequently, the same observed Stringy value can represent very different levels of unusualness at different sample sizes. For projection pursuit, where an optimizer may evaluate thousands of candidate projections, such sample-size dependence can distort how the index distinguishes structure from background noise.


Understanding this behaviour requires separating the ratio into its geometric components. For total Euclidean MST length, classical results provide a clear scaling law. In two dimensions and under suitable regularity conditions,

$$
L_n
$$

grows on the order of

$$
\sqrt n
$$
[@steele1988growth]. The weighted diameter is more difficult. Results exist for nearby quantities, such as the longest Euclidean MST edge [@penrose1997longest] and the graph diameter of random MSTs under different probabilistic models [@addarioberry2006diameter], but these results do not directly give the growth rate of the weighted Euclidean MST diameter used in Stringy.

This theoretical gap motivates the main investigation of the paper. Rather than assuming a growth rate for the numerator, we compare several candidate scaling functions empirically and examine which best removes systematic sample-size trends. Importantly, we do not judge a normalization only by whether it flattens the mean. A useful correction should also be examined in terms of relative spread and upper-tail behaviour, because these determine whether the statistic is genuinely becoming comparable across sample sizes.


The analysis also reveals a second problem. A deterministic correction based on separately scaling the numerator and denominator can substantially reduce the sample-size trend, but it no longer preserves the original unit-interval property

$$
0<S_n\leq1.
$$

This raises a broader methodological question: should the geometry of Stringy itself be redefined in order to remove sample-size effects, or should the original geometric index instead be retained and calibrated statistically against its null distribution?

The paper addresses five questions:

1. Does total Euclidean MST length behave as predicted by established Euclidean MST theory?
2. Can the weighted Stringy diameter be explained by nearby theoretical quantities such as the longest edge or hop diameter?
3. Which empirical scaling functions best stabilize the weighted diameter over the simulated range?
4. Does a scaling function that stabilizes the mean also stabilize relative dispersion and the upper tail?
5. Is sample-size-aware null calibration a more appropriate solution than replacing the original bounded definition of Stringy?

The paper is deliberately framed as an empirical scaling investigation rather than as a derivation of the asymptotic growth rate of the weighted Euclidean MST diameter. The contribution is instead to connect known MST theory with the finite-sample behaviour of Stringy, identify which component drives the sample-size dependence, evaluate the limits of deterministic scaling, and motivate a statistically calibrated version of the index for projection pursuit.

# Geometry of the Stringy statistic

Stringy is defined by a remarkably simple ratio, but that simplicity hides a more interesting geometric story.

Minimum spanning trees have long been used to summarize spatial relationships among observations. Classical statistical connections include single-linkage clustering [@gower1969mst] and later density-based cluster estimation using the sample MST [@stuetzle2003cluster]. In scagnostics, however, the role of the MST is different: it acts as a geometric skeleton of the scatterplot, from which visual characteristics of the point cloud can be quantified [@wilkinson2005graph; @wilkinson2008scagdist].

For Stringy, the key question is not simply how large the MST is. It is how much of that tree can be captured by one dominant path.

## Anatomy of the index

The distinction among total tree length, weighted diameter, and longest individual edge is easiest to understand by looking directly at the tree.

 @fig-stringy-anatomy shows the same structured point cloud and exactly the same MST, but highlights two different geometric objects. In the left panel, the emphasized path is the **weighted diameter**: the path through the MST whose accumulated Euclidean edge length is largest. This is the quantity used in the numerator of Stringy. In the right panel, only the **single longest edge** is highlighted.

The contrast is important. The longest edge is a local extreme: it identifies one unusually long connection in the tree. The weighted diameter is global: it can pass through many comparatively short edges and accumulate a much larger total length. Although both quantities come from the same MST, they answer fundamentally different geometric questions.

```{r}
#| label: fig-stringy-anatomy
#| fig-cap: "Two different geometric summaries of the same minimum spanning tree. Left: the weighted diameter, defined as the path with the greatest accumulated Euclidean length and used in the numerator of Stringy. Right: the single longest MST edge. The comparison illustrates why longest-edge theory does not directly determine the behaviour of the Stringy numerator."
#| fig-width: 9
#| fig-height: 4.5
set.seed(1050)
n_anatomy <- 200
u <- runif(n_anatomy, -1, 1)
polynomial <- poly(u, degree = 2)
x <- polynomial[, 1] + rnorm(n_anatomy, sd = 0.03)
y <- polynomial[, 2] + rnorm(n_anatomy, sd = 0.03)

sc_anatomy <- scree(x, y, out.rm = FALSE, binner = NULL)
mst_anatomy <- cassowaryr:::gen_mst(sc_anatomy$del, sc_anatomy$weights)
w_anatomy <- E(mst_anatomy)$weight

path_w <- get_diameter(mst_anatomy, directed = FALSE, weights = w_anatomy)
D_anatomy <- diameter(mst_anatomy, directed = FALSE, weights = w_anatomy)
L_anatomy <- sum(w_anatomy)
S_anatomy <- D_anatomy / L_anatomy

longest_edge <- E(mst_anatomy)[which.max(w_anatomy)]
longest_vertices <- ends(mst_anatomy, longest_edge, names = FALSE)

xy_anatomy <- sc_anatomy$del$x
path_segments <- data.frame(
  x = xy_anatomy[head(as.integer(path_w), -1), 1],
  y = xy_anatomy[head(as.integer(path_w), -1), 2],
  xend = xy_anatomy[tail(as.integer(path_w), -1), 1],
  yend = xy_anatomy[tail(as.integer(path_w), -1), 2]
)
edge_segment <- data.frame(
  x = xy_anatomy[longest_vertices[1], 1],
  y = xy_anatomy[longest_vertices[1], 2],
  xend = xy_anatomy[longest_vertices[2], 1],
  yend = xy_anatomy[longest_vertices[2], 2]
)

base_mst <- draw_mst(mst_anatomy, sc_anatomy) +
  coord_equal() + paper_theme +
  theme(axis.text = element_blank(), axis.ticks = element_blank())

p_anatomy_w <- base_mst +
  geom_segment(
    data = path_segments,
    aes(x, y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    colour = col_weighted,
    linewidth = 1.25
  ) +
  labs(
    title = "Weighted diameter path",
    subtitle = sprintf(
      "D = %.3f, L = %.3f, Stringy = %.3f",
      D_anatomy, L_anatomy, S_anatomy
    ),
    x = NULL,
    y = NULL
  )

p_anatomy_e <- base_mst +
  geom_segment(
    data = edge_segment,
    aes(x, y, xend = xend, yend = yend),
    inherit.aes = FALSE,
    colour = col_hop,
    linewidth = 1.5
  ) +
  labs(
    title = "Longest individual edge",
    subtitle = sprintf("E = %.3f", max(w_anatomy)),
    x = NULL,
    y = NULL
  )

p_anatomy_w + p_anatomy_e
```

The original Stringy index compares the length of the longest weighted path with the total length of the tree,

$$
S_n=\frac{D_n}{L_n}.
$$

Because every diameter path consists only of edges already contained in the MST,

$$
0<D_n\leq L_n,
$$

and therefore

$$
0<S_n\leq1.
$$

This bound is built directly into the geometry of the index. A value near one means that most of the tree can be traversed along a single dominant path; a smaller value indicates that more of the total tree length lies in branches away from that path.

This structural bound becomes especially important later. Any proposed normalization that modifies the numerator and denominator differently may reduce sample-size dependence, but it may also break the geometric relationship that guarantees Stringy remains between zero and one.

## Same distribution, increasingly dense MSTs

The sample-size problem becomes more intuitive when the tree is allowed to grow.

@fig-growing-mst shows four independent samples from exactly the same data-generating mechanism,

$$
(X,Y)\sim U(0,1)^2,
$$

with only the sample size changing. The square itself does not expand, the underlying distribution does not change, and no new structure is introduced. What changes is the resolution at which the region is sampled.

As $n$ increases, the MST fills the same spatial domain with more vertices and therefore more edges. Individual edges become shorter, but there are more of them. The total tree length, the weighted diameter, and their ratio do not respond to these changes in the same way.

```{r}
#| label: fig-growing-mst
#| fig-cap: "The geometry of the Euclidean MST as sampling density increases under the same Uniform(0,1)^2 data-generating mechanism. Grey segments show the complete MST and blue segments highlight the weighted diameter path. The spatial domain is unchanged across panels; only n increases. Each panel reports total MST length L, weighted diameter D, and Stringy S = D/L."
#| fig-width: 10
#| fig-height: 7
make_mst_panel <- function(n, seed) {
  set.seed(seed)
  x <- runif(n)
  y <- runif(n)
  sc <- scree(x, y, out.rm = FALSE, binner = NULL)
  mst <- cassowaryr:::gen_mst(sc$del, sc$weights)
  w <- E(mst)$weight
  path <- get_diameter(mst, weights = w)
  xy <- sc$del$x

  seg <- data.frame(
    x = xy[head(as.integer(path), -1), 1],
    y = xy[head(as.integer(path), -1), 2],
    xend = xy[tail(as.integer(path), -1), 1],
    yend = xy[tail(as.integer(path), -1), 2]
  )

  L <- sum(w)
  D <- diameter(mst, weights = w)

  draw_mst(mst, sc) +
    geom_segment(
      data = seg,
      aes(x, y, xend = xend, yend = yend),
      inherit.aes = FALSE,
      colour = col_weighted,
      linewidth = 0.8
    ) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    labs(
      title = paste0("n = ", n),
      subtitle = sprintf(
        "L = %.2f   D = %.2f   S = %.3f",
        L, D, D / L
      ),
      x = NULL,
      y = NULL
    ) +
    paper_theme +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      legend.position = "none"
    )
}

make_mst_panel(50, 51) +
  make_mst_panel(200, 52) +
  make_mst_panel(1000, 53) +
  make_mst_panel(3000, 54) +
  plot_layout(ncol = 2)
```


@fig-growing-mst shows why Stringy can change with sample size even when the underlying distribution does not change.

As $n$ increases, the MST contains more edges. These edges are generally shorter because the points are packed more densely, but there are many more of them. As a result, the **total MST length** can still increase.

The **weighted diameter** behaves differently because it uses only the edges along one longest path through the tree. When more points are added, this path may pass through more vertices, but the individual edges along the path also become shorter.

So the numerator and denominator of Stringy respond to increasing sample size in different ways.

The key question is therefore not simply whether Stringy changes with $n$, but **why** it changes. In particular, we want to know whether the total tree length $L_n$ and the longest weighted path $D_n$ grow at different rates.

The rest of this paper investigates this difference.




# References
